# Executing `python` code {#sec-python}

::: {.callout-warning icon='true'}

## Important points

- `BiocBook`s can execute **`python` code**, not just `R` code;
- The code runs on **every render**, including when the *Bioconductor Build System* 
  builds the book;
- The `python` packages the book needs are therefore installed **at build time**,
  by the book itself, from the `conda` environment declared in `inst/requirements.yml`;
- **Keep at least one `R` chunk on a `python` page**: it is what keeps `quarto` on 
  the `knitr` engine.

:::

## Declaring and installing the python environment

`python` packages are listed, pinned, in `inst/requirements.yml` -- a plain
`conda` environment file, e.g.

``` yaml
name:
    BiocBook
channels:
    - conda-forge
    - bioconda
    - nodefaults
dependencies:
    - python=3.12
    - numpy=1.26
    - matplotlib
```

Note the presence of the `nodefaults` channel: the `defaults` channel is covered by the Anaconda Terms
of Service, which require a paid licence for many organisations. `conda-forge`
and `bioconda` are community channels that are free to use.

The first page that needs `python` installs and activates them:

```{r "setup-python"}
#| include: true
library(reticulate)
BiocBook::setup_python()
```

`setup_python()` finds `inst/requirements.yml`,  creates the `conda` environment 
declared there (if not installed yet) and activates it, 
addressing it by full path rather than by name. 
If the environment already exists, it is re-used rather than resolved again, 
and if `RETICULATE_PYTHON` is already set (as it is inside the book's `Docker` 
image) the whole step short-circuits.

## Where `conda` itself comes from

Building a `conda` environment needs a `conda` implementation, and the machines
that build this book do not necessarily have one. Bioconductor's current builders
may, and the `r-universe` build image ships `quarto`, `python3` and `pip`, 
but no `conda` at all. So `BiocBook` brings its own. 

`micromamba` is a single self-contained binary: no `conda` installation, 
no base environment, no `python`. 

`BiocBook::micromamba()` resolves it in order:

1. `RETICULATE_CONDA`, if you have set it;
2. a `micromamba` already on the `PATH` (this book's `Docker` image installs one);
3. a copy previously downloaded into `BiocBook`'s cache;
4. failing all of those, it downloads a **pinned** release and checks it against
   a recorded `SHA256` before using it.

::: {.callout-note icon='true'}

## Keep an `R` chunk on the page

`quarto` chooses its execution engine **per file**. A page with at least one
`R` chunk uses `knitr`, and runs its `python` chunks through `reticulate`. A
page whose only code is `python` uses `jupyter` instead. However, the Bioconductor 
builders do not provide `jupyter`, so a `jupyter` page cannot be rendered there at all.

To force `quarto` to use `knitr`, keep at least one `R` chunk on the page, even 
if it is empty.

:::

## A worked example

`reticulate` hosts a single, persistent `python` session that is shared with
`R`, so objects can be passed in both directions.

Start from an `R` object:

```{r "r-object"}
counts <- matrix(
    c(12, 0, 45, 3, 7, 91, 0, 5, 33),
    nrow = 3,
    dimnames = list(c("geneA", "geneB", "geneC"), c("s1", "s2", "s3"))
)
counts
```

The `R` object is reachable from `python` through the `r` object:

```{python "python-reads-r"}
import numpy as np
counts = np.array(r.counts)
print("shape:", counts.shape)
print("library sizes:", counts.sum(axis = 0))
```

State persists from one `python` chunk to the next, exactly as it would in a
notebook:

```{python "python-state"}
cpm = counts / counts.sum(axis = 0) * 1e6
print("CPM values:")
print(np.round(cpm, 1))
```

Finally, the `python` object can be passed back to `R`:

```{r "r-reads-python"}
cpm <- reticulate::py$cpm
cpm
```

## Bioconda packages 

Because the `BiocBook` environment is a `conda` environment, it can install any package
from the `bioconda` channel, e.g. `deeptools`:

```{python "python-deeptools"}
from importlib.metadata import version
v = version("deeptools")
```

Then read that in `R`:

```{r "r-reads-deeptools"}
reticulate::py$v
```

## Session info

```{r}
sessioninfo::session_info()
```
