---
eval: false
---

# Containerizing and Publishing against specific Bioconductor release versions

## For a {`BiocBook`} package not currently in Bioconductor

One can build against older Bioconductor releases as follows: 

- Commit current changes before switching branches (should be in `devel` branch)

```{r}
gert::git_commit_all("Commit current changes")
```

- Create a new branch named `RELEASE_X_Y` and checkout

```{r}
gert::git_branch_create("RELEASE_3_17")
```

- Push local `RELEASE_X_Y` to Github

```{r}
gert::git_push()
```

## For a {`BiocBook`} package accepted in Bioconductor > 6 months ago

Once your package is accepted in Bioconductor, your repository will have 
access to a new **remote** named `upstream`, pointing to `git@git.bioconductor.org`. 
When Bioconductor releases a new version, your package will change version on 
the `upstream` remote, in a dedicated release branch `RELEASE_X_Y`. 
All details are available [here](https://contributions.bioconductor.org/git-version-control.html). 

To generate a Docker image and a version of the `BiocBook` website built 
on new Bioconductor releases, you can: 

- Add the `Bioconductor` remote to your local repository (this might already be set up)

```{r}
gert::git_remote_add(name = 'upstream', url = 'git@git.bioconductor.org:packages/<YOUR-REPOSITORY>.git')
```

- Create a new local `RELEASE_X_Y` branch 

```{r}
gert::git_branch_create("RELEASE_3_15")
```

- Pull the `upstream` `RELEASE_X_Y` commits

```{r}
gert::git_pull("RELEASE_3_15", remote = "upstream")
```

- Push the local `RELEASE_X_Y` branch to `origin` remote (your own Github repository)

```{r}
gert::git_push("RELEASE_3_15", remote = "origin")
```

## Serving the book to AI assistants

Each version of a `BiocBook` also serves an `llms.txt` file (e.g. 
[`devel/llms.txt`](https://js2264.github.io/BiocBookDemo/devel/llms.txt)), 
following the [llms.txt proposal](https://llmstxt.org/): a Markdown index of 
the book that AI assistants can read instead of crawling its HTML pages. It 
links to a Markdown copy of every page, e.g. `pages/Chapter-4.llms.md` next to 
`pages/Chapter-4.html`.

- `quarto` >= 1.11 writes these files when the book's `llms-txt` option is on. 
  Books created from template >= 1.1.0 switch it on through the `llms` `quarto` 
  profile (`inst/_quarto-llms.yml`), which `vignettes/Makefile` activates 
  whenever the installed `quarto` supports it (older versions reject the 
  option). The `biocbook` GitHub workflow pins such a `quarto` with its 
  `QUARTO_VERSION` setting, and copies built by Bioconductor get these files 
  once the Bioconductor builders run `quarto` >= 1.11.
- After every render, `inst/assets/post-render.R` calls 
  `BiocBook::enrich_llms_txt()`, which adds what an assistant needs to run the 
  book's code to `llms.txt`: the book package and its version, how to install 
  it, its `Docker` image, its `python` environment and where the other 
  versions of the book live. With an older `quarto`, it writes a minimal 
  `llms.txt` listing the chapters.

To render these files locally: 

```{sh filename="bash"}
QUARTO_PROFILE=llms quarto render inst/
```

## Session info

```{r}
sessioninfo::session_info()
```
